---
title: "Campaign Effectiveness Analysis: Measuring Incremental Revenue with a Geo Experiment"
author: "Jean Liu"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

# Business Context

A retailer launched a regional marketing campaign to evaluate whether advertising generated incremental revenue.

The campaign was tested through a geo-based experiment using Designated Market Areas (DMAs). Some markets were assigned to a treatment group, while others served as a control group.

This analysis estimates the causal impact of the campaign and evaluates whether the marketing investment should be scaled.

# Business Question

Did the advertising campaign generate incremental revenue? 

More specifically,

- Was revenue growth driven by advertising?
- Or were treatment markets already larger before the campaign?

Answering this question is critical for determining whether the campaign should be scaled, modified, or discontinued.

## Executive Summary

```{r}
library(knitr)

executive_summary <- tibble(
  `Key Findings` = c(
    "No statistically significant incremental lift was detected.",
    "No significant pre-campaign difference was detected between treatment and control markets.",
    "Recommendation"
  ),
  Business Interpretation = c(
    "The campaign did not generate measurable incremental revenue during the test period.",
    "The experimental groups were sufficiently balanced for causal analysis.",
    "Do not scale the campaign based on the current evidence. Consider testing alternative targeting strategies, creative assets, or media channels before additional investment."
  )
)

kable(
  executive_summary,
  caption = "Executive Summary"
)
```

# Experimental Design

The campaign was implemented as a geo experiment.

Selected Designated Market Areas (DMAs) received advertising, while comparable DMAs served as a control group.

Revenue was observed before and after the campaign.

The analysis estimates advertising incrementality by comparing revenue changes across treatment and control markets rather than relying on raw revenue differences.

> Analysis workflow

Geo Experiment
      ↓
Data Preparation
      ↓
Baseline Assessment
      ↓
Naive Before–After Analysis
      ↓
Incrementality Estimation (DiD)
      ↓
Business Recommendation

# Data Preparation

```{r}
library(tidyverse)

campaign <- read_csv("data/Date of Advertising.csv")
```

```{r}
glimpse(campaign)
summary(campaign)
```

```{r}
campaign <- campaign |>
  mutate(
    date = as.Date(date, format = "%m/%d/%Y"),

    period = factor(
      isTreatmentPeriod,
      levels = c(0, 1),
      labels = c("Before", "After")
    ),

    group = factor(
      isTreatmentGroup,
      levels = c(0, 1),
      labels = c("Control", "Treatment")
    )
  )

glimpse(campaign)


summary_table <- campaign |>
  group_by(group, period) |>
  summarise(
    n_observations = n(),
    n_dmas = n_distinct(dma),
    avg_revenue = mean(revenue),
    median_revenue = median(revenue),
    total_revenue = sum(revenue),
    .groups = "drop"
  )

summary_table
```

The dataset contains observations from both treatment and control markets across pre- and post-campaign periods, providing the foundation for estimating incremental advertising impact.

# Baseline Comparison

Before estimating campaign impact, we need to check whether treatment and control markets were comparable before the campaign.

```{r}
ggplot(pre_period,
       aes(x = log(revenue),
           fill = group)) +
  geom_density(alpha = 0.4) +
  labs(
    title = "Distribution of Pre-Campaign Log Revenue",
    subtitle = "Treatment and control markets exhibit similar baseline distributions",
    x = "Log Revenue",
    y = "Density",
    fill = "Group"
  )
```

Although treatment markets exhibited higher average revenue in raw values, the distribution of log-transformed revenue is broadly similar across groups. This observation is consistent with the formal balance check presented below.

Because raw revenue levels differ across markets, comparing post-campaign revenue alone could lead to misleading conclusions. A formal pre-campaign balance check and a Difference-in-Differences model are therefore used to isolate the incremental effect of advertising.

# Pre-Campaign Balance Check

Before estimating campaign impact, we first check whether treatment and control markets were statistically similar during the pre-campaign period.

```{r}
pre_period <- campaign |>
    filter(period=="Before")
```

```{r}
randomization_model <- lm(
  log(revenue) ~ group,
  data = pre
)

summary(randomization_model)
```

Before estimating campaign impact, we verify whether treatment assignment produced comparable markets before the intervention.

p-value = 0.5698, No statistically significant difference was observed before the campaign, supporting the validity of the experimental design.

# Naive Before–After Estimate

A common way to evaluate a marketing campaign is to compare revenue before and after the campaign within the treatment markets.

However, this approach assumes that all observed changes are caused by the campaign, ignoring other factors such as seasonality, macroeconomic conditions, and overall market trends.

The following regression estimates the average revenue difference before and after the campaign for treatment markets only.

```{r}
treatment <- campaign |>
  filter(group == "Treatment")

before_after_model <- lm(
  log(revenue) ~ period,
  data = treatment
)

summary(before_after_model)
```

The estimated coefficient measures the average revenue change after the campaign within treatment markets.

Although this estimate may suggest that revenue changed following the campaign, it cannot be interpreted as the causal effect of advertising because no comparison is made with markets that were not exposed to the campaign.

Consequently, this approach may attribute unrelated market-wide changes to the advertising campaign, leading to biased estimates of marketing effectiveness.

# Incrementality Estimation

```{r}
campaign_trend <- campaign |>
  group_by(group, period) |>
  summarise(
    avg_revenue = mean(revenue),
    .groups = "drop"
  )

ggplot(campaign_trend,
       aes(x = period,
           y = avg_revenue,
           group = group,
           color = group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  labs(
    title = "Revenue Trend Before and After Campaign Launch",
    subtitle = "Both groups experienced similar revenue trends over time",
    x = NULL,
    y = "Average Revenue",
    color = "Group"
  )
```

The figure suggests that treatment and control markets followed similar revenue trends over time. To formally estimate whether the campaign generated incremental revenue beyond these common trends, we apply a Difference-in-Differences (DiD) model.

```{r}
incrementality_model <- lm( 
    log(revenue) ~ isTreatmentGroup * isTreatmentPeriod, 
    data = campaign 
) 
```

```{r}
library(broom)
library(dplyr)
library(knitr)

results <- tidy(incrementality_model) |>
  filter(term != "(Intercept)") |>
  mutate(
    Variable = c(
      "Treatment",
      "Post Period",
      "Treatment × Post"
    ),
    Interpretation = c(
      "No significant baseline difference",
      "Overall revenue declined slightly",
      "No significant incremental advertising effect"
    )
  ) |>
  transmute(
    Variable,
    Estimate = round(estimate, 3),
    `P-value` = round(p.value, 3),
    Interpretation
  )

kable(
  results,
  caption = "Difference-in-Differences Regression Results"
)
```

The trend plot suggests that treatment and control markets experienced similar revenue changes over time. The Difference-in-Differences model formally estimates whether the treatment markets experienced additional revenue growth beyond this common trend. The interaction term captures this incremental advertising effect.

# Business Recommendation

The Difference-in-Differences analysis found no statistically significant evidence that the advertising campaign generated incremental revenue (interaction coefficient = 0.0066, p = 0.845).

Based on this experiment alone, there is insufficient evidence to justify scaling the campaign to additional markets.

Future iterations should consider testing alternative targeting strategies, creative content, or channel allocation before expanding media spend.

- **Decision**: Do not expand campaign investment based on the current experiment.
- **Risk**: Scaling an ineffective campaign could increase marketing spend without generating incremental revenue.
- **Next Step**: Conduct follow-up experiments with alternative targeting, creative assets, or media allocation.

# Limitations

This analysis is based on a geo-level experiment, so the results may not generalize to all advertising campaigns.

Potential limitations include market-level differences, seasonality, regional demand shocks, and the fact that results from one campaign or channel may not apply to other media strategies.