---
title: "Campaign Effectiveness Analysis: Measuring Incremental Revenue with a Geo Experiment"
author: "Jean Liu"

format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-summary: "Show Code"
    theme: cosmo

execute:
  warning: false
  message: false
---

# Business Context

A retailer launched a regional marketing campaign to evaluate whether advertising generated incremental revenue.

The campaign was tested through a geo-based experiment using Designated Market Areas (DMAs). Some markets were assigned to a treatment group, while others served as a control group.

This analysis estimates the causal impact of the campaign and evaluates whether the marketing investment should be scaled.

# Business Question

Did the advertising campaign generate incremental revenue? 

More specifically,

- Was revenue growth driven by advertising?
- Or were treatment markets already larger before the campaign?

Answering this question is critical for determining whether the campaign should be scaled, modified, or discontinued.

## Executive Summary

```{r}
library(tidyverse)
library(knitr)
library(tibble)
library(broom)

executive_summary <- tibble(
  Finding = c(
    "No Incremental Lift",
    "Balanced Experiment",
    "Recommendation"
  ),
  Evidence = c(
    "DiD interaction p = 0.845",
    "Pre-campaign balance p = 0.570",
    "No statistically significant campaign effect"
  ),
  `Business Impact` = c(
    "Advertising did not generate measurable incremental revenue.",
    "Treatment and control markets were comparable before the campaign.",
    "Do not scale the campaign; test alternative targeting, creatives, or media channels."
  )
)

kable(
  executive_summary,
  caption = "Executive Summary"
)
```

# Experimental Design

The campaign was implemented as a geo experiment.

Selected Designated Market Areas (DMAs) received advertising, while comparable DMAs served as a control group.

Revenue was observed before and after the campaign.

The analysis estimates advertising incrementality by comparing revenue changes across treatment and control markets rather than relying on raw revenue differences.

> Analysis workflow

```{mermaid}
flowchart LR
    A[Geo Experiment]
    B[Data Preparation]
    C[Baseline Comparison]
    D[Naive Before-After]
    E[Incrementality Estimation]
    F[Business Recommendation]

    A --> B --> C --> D --> E --> F
```

# Data Preparation

```{r}
campaign <- read_csv("../data/Date of Advertising.csv")

campaign <- campaign |>
  mutate(
    date = as.Date(date, format = "%m/%d/%Y"),

    period = factor(
      isTreatmentPeriod,
      levels = c(0, 1),
      labels = c("Before", "After")
    ),

    group = factor(
      isTreatmentGroup,
      levels = c(0, 1),
      labels = c("Control", "Treatment")
    )
  )

summary_table <- campaign |>
  group_by(group, period) |>
  summarise(
    n_observations = n(),
    n_dmas = n_distinct(dma),
    avg_revenue = mean(revenue),
    median_revenue = median(revenue),
    total_revenue = sum(revenue),
    .groups = "drop"
  )

summary_table |>
  rename(
    Group = group,
    Period = period,
    Observations = n_observations,
    DMAs = n_dmas,
    `Average Revenue` = avg_revenue,
    `Median Revenue` = median_revenue,
    `Total Revenue` = total_revenue
  ) |>
  kable(
    digits = 0,
    caption = "Campaign Summary by Group and Period"
  )
```

The dataset contains observations from both treatment and control markets across pre- and post-campaign periods, providing the foundation for estimating incremental advertising impact.

# Baseline Comparison

Before estimating campaign impact, we need to check whether treatment and control markets were comparable before the campaign.

```{r}
pre_period <- campaign |>
  filter(period == "Before")

ggplot(
  pre_period,
  aes(
    x = log(revenue),
    fill = group
  )
) +
  geom_density(alpha = 0.35) +
  scale_fill_manual(
    values = c(
      "Control" = "#9AA0A6",
      "Treatment" = "#2563EB"
    )
  ) +
  labs(
    title = "Pre-Campaign Revenue Distribution",
    subtitle = "Treatment and control markets showed similar baseline distributions",
    x = "Log Revenue",
    y = "Density",
    fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray40")
  )
```

Although treatment markets exhibited higher average revenue in raw values, the distribution of log-transformed revenue is broadly similar across groups. This observation is consistent with the formal balance check presented below.

Because raw revenue levels differ across markets, comparing post-campaign revenue alone could lead to misleading conclusions. A formal pre-campaign balance check and a Difference-in-Differences model are therefore used to isolate the incremental effect of advertising.

# Pre-Campaign Balance Check

Before estimating campaign impact, we first check whether treatment and control markets were statistically similar during the pre-campaign period.

```{r}
randomization_model <- lm(
  log(revenue) ~ group,
  data = pre_period
)

tidy(randomization_model) |>
  filter(term != "(Intercept)") |>
  transmute(
    Variable = "Treatment Group",
    Estimate = round(estimate, 3),
    `P-value` = round(p.value, 3)
  ) |>
  kable(
    caption = "Pre-Campaign Balance Check"
  )
```

No statistically significant baseline difference was detected between treatment and control markets (p = 0.570), supporting the validity of the experimental design.

# Naive Before–After Estimate

A common way to evaluate a marketing campaign is to compare revenue before and after the campaign within the treatment markets.

However, this approach assumes that all observed changes are caused by the campaign, ignoring other factors such as seasonality, macroeconomic conditions, and overall market trends.

The following regression estimates the average revenue difference before and after the campaign for treatment markets only.

```{r}
treatment <- campaign |>
  filter(group == "Treatment")

before_after_model <- lm(
  log(revenue) ~ period,
  data = treatment
)

tidy(before_after_model) |>
  filter(term != "(Intercept)") |>
  transmute(
    Variable = "After Period",
    Estimate = round(estimate, 3),
    `P-value` = round(p.value, 3),
    Interpretation = "Revenue declined within treatment markets"
  ) |>
  kable(
    caption = "Naive Before–After Regression Results"
  )
```

Although revenue declined after the campaign within treatment markets (β = -0.039, p = 0.047), this estimate cannot be interpreted as the causal effect of advertising because it does not control for broader market trends.

# Incrementality Estimation

```{r}
campaign_trend <- campaign |>
  group_by(group, period) |>
  summarise(
    avg_revenue = mean(revenue),
    .groups = "drop"
  )

ggplot(
  campaign_trend,
  aes(
    x = period,
    y = avg_revenue,
    group = group,
    color = group
  )
) +
  geom_line(linewidth = 1.4) +
  geom_point(size = 4) +
  scale_color_manual(
    values = c(
      "Control" = "#9AA0A6",
      "Treatment" = "#2563EB"
    )
  ) +
  scale_y_continuous(
    labels = scales::comma,
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title = "Revenue Trends by Market",
    subtitle = "Both groups declined at a similar rate after the campaign",
    x = NULL,
    y = "Average Revenue ($)",
    color = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray40")
  )
```

The figure suggests that treatment and control markets followed similar revenue trends over time. To formally estimate whether the campaign generated incremental revenue beyond these common trends, we apply a Difference-in-Differences (DiD) model.

```{r}

incrementality_model <- lm(
  log(revenue) ~ isTreatmentGroup * isTreatmentPeriod,
  data = campaign
)

results <- tidy(incrementality_model) |>
  filter(term != "(Intercept)") |>
  mutate(
    Variable = c(
      "Treatment",
      "Post Period",
      "Treatment × Post"
    ),
    Interpretation = c(
      "No significant baseline difference",
      "Overall revenue declined slightly",
      "No significant incremental advertising effect"
    )
  ) |>
  transmute(
    Variable,
    Estimate = round(estimate, 3),
    `P-value` = round(p.value, 3),
    Interpretation
  )

kable(
  results,
  caption = "Difference-in-Differences Regression Results"
)
```

The trend plot suggests that treatment and control markets experienced similar revenue changes over time. The Difference-in-Differences model formally estimates whether the treatment markets experienced additional revenue growth beyond this common trend. The interaction term captures this incremental advertising effect.

# Business Recommendation

The Difference-in-Differences model found no statistically significant incremental advertising effect (β = 0.007, p = 0.845). Based on the current evidence, campaign expansion is not recommended.

```{r}
recommendation_table <- tibble(
  Decision = c(
    "Do not scale the campaign",
    "Optimize targeting",
    "Test new creatives or media channels"
  ),
  Rationale = c(
    "No statistically significant incremental lift was detected.",
    "Current treatment markets did not show additional growth versus control markets.",
    "The current campaign setup may not be strong enough to generate measurable lift."
  ),
  `Recommended Action` = c(
    "Avoid expanding media spend based on this experiment alone.",
    "Run a follow-up geo experiment with refined audience segments.",
    "Evaluate alternative creative assets, messaging, or channel allocation."
  )
)

kable(
  recommendation_table,
  caption = "Business Recommendation"
)
```

Overall, the current evidence does not support scaling the campaign. Future investment should be tied to follow-up experiments with clearer targeting, creative variation, and measurable lift criteria.

# Limitations

This analysis is based on a geo-level experiment, so the results may not generalize to all advertising campaigns.

Potential limitations include market-level differences, seasonality, regional demand shocks, and the fact that results from one campaign or channel may not apply to other media strategies.