---
title: "Facebook Advertising Performance Analysis"
author: "Jean Liu"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

# Project Overview

This project analyzes Facebook advertising performance for retail brands using data from a large-scale advertising experiment.

The business objective is to identify which ad attributes are associated with higher click efficiency, measured by clicks per dollar.

The analysis focuses on:

- Campaign performance overview
- Creative and placement performance
- Regression-based performance driver analysis
- Audience interaction effects
- Business recommendations

```{r}
library(tidyverse)
library(broom)
library(scales)

facebook <- read_csv("data/Facebook Data.csv")

glimpse(facebook)
```

```{r}
names(facebook)
```

```{r}
facebook_clean <- facebook |>
  select(-`...1`) |>
  mutate(
    date = as.Date(date, "%m/%d/%Y"),

    adType = factor(adType),
    placement = factor(placement),
    category = factor(category),
    keywords = factor(keywords),
    body = factor(body)
  )

# Check data structure
glimpse(facebook_clean)

```

# Data Audit

Before analyzing advertising performance, the dataset was reviewed to understand its structure, variable types, and overall quality.

This step verifies that the data are suitable for subsequent exploratory analysis and regression modeling.

```{r}
dim(facebook_clean)

summary(facebook_clean)

colSums(is.na(facebook_clean))
```

The dataset contains **4,431 Facebook advertisements** across six retail categories.

After data cleaning, all variables were successfully converted to appropriate data types, and no missing values were detected, indicating that the dataset is suitable for exploratory analysis and regression modeling.

# Campaign Overview

This section summarizes how advertisements were distributed across placements, ad types, categories, and audience age groups.

```{r}
facebook_clean |>
  count(placement)

facebook_clean |>
  count(adType)

facebook_clean |>
  count(category)
```

# Exploratory Data Analysis

Before estimating regression models, exploratory analysis was conducted to understand the distribution of advertising performance and identify patterns across campaign attributes.

```{r}
ggplot(facebook_clean,
       aes(clickPerDollar)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Clicks per Dollar",
    x = "Clicks per Dollar",
    y = "Count"
  )
```

Clicks per Dollar appears approximately continuous without severe outliers, suggesting that linear regression is an appropriate starting point for modeling advertising performance.

```{r}
ggplot(facebook_clean,
       aes(ageMean)) +
  geom_histogram(bins = 20) +
  labs(
    title = "Distribution of Target Audience Age",
    x = "Average Audience Age",
    y = "Count"
  )
```

```{r}
ggplot(facebook_clean,
       aes(
         x = placement,
         y = clickPerDollar
       )) +
  geom_boxplot() +
  labs(
    title = "Advertising Performance by Placement",
    x = "Placement",
    y = "Clicks per Dollar"
  )
```

Advertisements shown on mobile devices generally achieved higher click efficiency than desktop advertisements, although some variation exists within both groups.

```{r}
ggplot(facebook_clean,
       aes(
         x = adType,
         y = clickPerDollar
       )) +
  geom_boxplot() +
  labs(
    title = "Advertising Performance by Ad Type",
    x = "Ad Type",
    y = "Clicks per Dollar"
  )
```

Photo posts generally outperformed link posts, suggesting that visual advertising formats may generate stronger user engagement.

```{r}
ggplot(facebook_clean,
       aes(
         x = category,
         y = clickPerDollar
       )) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(
    title = "Advertising Performance Across Retail Categories",
    x = "Retail Category",
    y = "Clicks per Dollar"
  )
```

Advertising performance varies across retail categories, indicating that campaign effectiveness may depend on product type and customer interest.

```{r}
ggplot(facebook_clean,
       aes(ageMean,
           clickPerDollar)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  labs(
    title = "Audience Age and Advertising Performance",
    x = "Average Audience Age",
    y = "Clicks per Dollar"
  )
```

# Performance Summary

This section compares average click efficiency across placements, ad types, retail categories, and creative messages.

The goal is to identify which campaign attributes are associated with stronger advertising performance before moving into regression modeling.

```{r}
facebook_clean |>
  summarise(
    avg_click_per_dollar = mean(clickPerDollar, na.rm = TRUE),
    median_click_per_dollar = median(clickPerDollar, na.rm = TRUE),
    sd_click_per_dollar = sd(clickPerDollar, na.rm = TRUE),
    avg_ctr = mean(clickThroughRate, na.rm = TRUE)
  )
```

```{r}
facebook_clean |>
  group_by(placement) |>
  summarise(
    avg_click_per_dollar = mean(clickPerDollar, na.rm = TRUE),
    avg_ctr = mean(clickThroughRate, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  arrange(desc(avg_click_per_dollar))
```

Mobile advertisements generated substantially higher click efficiency than desktop advertisements, suggesting that mobile placement may provide stronger advertising performance.

```{r}
facebook_clean |>
  group_by(adType) |>
  summarise(
    avg_click_per_dollar = mean(clickPerDollar, na.rm = TRUE),
    avg_ctr = mean(clickThroughRate, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  arrange(desc(avg_click_per_dollar))
```

Photo posts outperformed link posts in both click-through rate and clicks per dollar, indicating that visual creatives may be more effective in driving user engagement.

```{r}
facebook_clean |>
  group_by(category) |>
  summarise(
    avg_click_per_dollar = mean(clickPerDollar, na.rm = TRUE),
    avg_ctr = mean(clickThroughRate, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  arrange(desc(avg_click_per_dollar))
```

Among all retail categories, Cosmetics achieved the highest click efficiency, while eCommerce and General retail categories produced comparatively lower performance.

```{r}
facebook_clean |>
  group_by(body) |>
  summarise(
    avg_click_per_dollar = mean(clickPerDollar, na.rm = TRUE),
    avg_ctr = mean(clickThroughRate, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  arrange(desc(avg_click_per_dollar))
```

Creative messages encouraging users to "Like" the brand consistently achieved the highest click efficiency, suggesting that direct engagement-oriented messaging performs better than more generic promotional copy.

The descriptive results provide an initial view of which placements, ad types, categories, and creative messages appear to perform better. However, these comparisons do not control for other campaign attributes, so regression modeling is needed to estimate performance drivers more carefully.

# Regression Modeling

Descriptive statistics provide useful comparisons across campaign attributes, but they cannot isolate the independent effect of each advertising characteristic.

To better understand which factors are associated with advertising performance, a multiple linear regression model is estimated while controlling for other campaign attributes.

```{r}
model_1 <- lm(
  clickPerDollar ~
    adType +
    placement +
    keywords +
    body +
    category +
    ageMean,
  data = facebook_clean
)

summary(model_1)
```

```{r}
library(broom)

tidy(model_1)
```

```{r}
glance(model_1)
```

The model explains approximately **88.7%** of the variation in clicks per dollar, indicating strong explanatory power for observed campaign performance.

## Significant Performance Drivers

Key evaluation metrics are reported below.

```{r}

significant_results <- model1_results |>
  filter(
    term != "(Intercept)",
    p.value < 0.05
  )

ggplot(
  significant_results,
  aes(
    x = estimate,
    y = reorder(term, estimate)
  )
) +
  geom_point() +
  geom_vline(
    xintercept = 0,
    linetype = "dashed"
  ) +
  labs(
    title = "Significant Drivers of Advertising Performance",
    x = "Estimated Effect on Clicks per Dollar",
    y = NULL
  )
```

```{r}
model_2 <- lm(
  clickPerDollar ~
    adType +
    placement +
    keywords +
    body +
    category * ageMean,
  data = facebook_clean
)

summary(model_2)

model2_results <- tidy(model_2)

model2_results

glance(model_2)
```

```{r}
ggplot(facebook_clean,
       aes(x = ageMean,
           y = clickPerDollar,
           color = category)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Interaction Between Audience Age and Retail Category",
    x = "Average Audience Age",
    y = "Clicks per Dollar",
    color = "Category"
  )
```

The exploratory visualization suggests that the relationship between audience age and advertising performance may vary across retail categories. This observation is formally evaluated using the interaction regression model in the following section.

# Interaction Analysis

A second regression model introduces an interaction between retail category and audience age to examine whether the relationship between age and advertising performance differs across product categories.

This analysis evaluates whether age has a consistent effect across all retail categories or whether campaign effectiveness varies depending on the type of retailer.

# Model Comparison

```{r}
anova(model_1, model_2)

glance(model_1)
glance(model_2)
```

The interaction model provides a statistically significant improvement over the baseline model (ANOVA p < 0.001).

Model fit also improves slightly, with the adjusted R² increasing from **0.886** to **0.890**, indicating that the relationship between audience age and advertising performance differs across retail categories.

# Key Findings

The analysis identified several consistent drivers of Facebook advertising performance.

- Mobile placement generated substantially higher click efficiency than desktop placement.
- Photo posts outperformed link posts, suggesting that visual creatives attract stronger engagement.
- Engagement-focused messaging encouraging users to "Like" the brand consistently achieved the highest click efficiency.
- Campaign performance varied across retail categories.
- The interaction model indicates that audience age influences advertising performance differently across retail categories.

# Business Recommendations

Based on the analysis, the following recommendations are proposed to improve Facebook advertising performance:

- Prioritize mobile placements when allocating advertising budgets.
- Increase the use of photo-based creatives over link posts.
- Continue testing engagement-oriented messaging that encourages users to interact with the brand.
- Tailor campaign strategies according to retail category rather than adopting a uniform approach.
- Consider audience demographics when designing campaigns, as the effectiveness of age targeting varies across product categories.